
## Basit komutlar

Örneğin, bir sayı vektörü oluşturmak için c()( concatenate için ) işlevini kullanırız.
Parantez içindeki tüm sayılar birleştirilir. Aşağıdaki komut R, 1, 3, 2 ve 5 sayılarını birleştirmeyi ve bunları x. 
Yazdığımızda x, bize vektörü geri verir.

```{r chunk1}
x <- c(1, 3, 2, 5)
x
```

`<-` bu işaretin yerine "=" işaretide ayn atama işlemini yapar.

```{r chunk2}
x = c(1, 6, 2)
x
y = c(1, 4, 3)
```

R 'ye İki sayı kümesini birbirine eklemeyi söyleyebiliriz . x ve y vektörlerini toplayalım.

Ancak, x ve y aynı uzunlukta olmalıdır.

length() Fonksiyonu kullanarak uzunluklarını kontrol edebiliriz .

```{r chunk3}
length(x)
length(y)
x + y
```


ls(), şimdiye kadar kaydettiğimiz veriler ve işlevler gibi tüm nesnelerin listesine bakmamızı sağlar. 

rm(), istemediğimiz herhangi bir şeyi silmek için kullanılabilir.


```{r chunk4}
ls()
rm(x, y)
ls()
```

Aynı anda tüm nesneleri kaldırmak da mümkündür:

```{r chunk5}
rm(list = ls())
```


matrix(), bir sayı matrisi oluşturmak için kullanılabilir. İşlevi kullanmadan önce matrix(), onun hakkında daha fazla bilgi edinebiliriz:


```{r chunk6}
?matrix
```


Yardım dosyası, matrix()işlevin bir dizi girdi aldığını gösterir, ancak şimdilik ilk üçüne odaklanıyoruz: veriler (matristeki girişler), satır sayısı ve sütun sayısı. İlk önce basit bir matris oluşturuyoruz.


```{r chunk7}
x = matrix(data = c(1,2,3,4), nrow = 2, ncol = 2)
x
```

Yukarıdaki komutta "data", "nrow" ve "ncol" yazmayı atlayabileceğimize dikkat.


```{r chunk8}

x = matrix(c(1, 2, 3, 4), 2, 2)
x
```
byrow işlevi de sayları satır boyunca dizmeye yarar yuardaki örnekte süütüna göre dizmişti.


```{r chunk9}
matrix(c(1, 2, 3, 4), 2, 2, byrow = TRUE)
```


sqrt()` funksiyonu her elemann kare köklerini döndürür. The command



```{r chunk10}
sqrt(x)
x^2
```

`rnorm()`normal dağılmış değişkenler oluşturur özellikleri : ortalams(mean) 0 olan ve standart sapmas 1 olan bir vektör tanmlar.Ama yinede bu ortalama ve standart sapmay değiştirebiliriz.

y diye bir değişken tanplayp ortalamsı 50 olan ve standart sapması 0.1 olan bir vektör tanımlayalım.


```{r chunk11}
x = rnorm(50)

y <- x + rnorm(50, mean = 50, sd = .1)
y
cor(x, y)

x

```


Eğer rasgele üretilen saylarn her zaman aynı sayları vermesini istediğimizde set.seed() fonksiyonunu kullanırız.


```{r chunk12}
set.seed(1303)
rnorm(50)
```

Mean() ve var() işlevleri, bir sayı vektörünün ortalamasını ve varyansını hesaplamak için kullanılabilir.

var() çıktısına sqrt() uygulamak standart sapmayı verecektir. Ya da basitçe sd() işlevini kullanabiliriz.

```{r chunk13}
set.seed(3)
y = rnorm(100)
mean(y)
var(y)
sqrt(var(y))
sd(y)
```


## Graphics
plot(), verileri çizmenin birincil yoludur.


```{r chunk14}
x <- rnorm(100)
y <- rnorm(100)
plot(x, y)
plot(x, y, xlab = "this is the x-axis",
    ylab = "this is the y-axis",
    main = "Plot of X vs Y")
```

 `pdf()` fonksiyonu ile yukarıdaki grafiği kaydedebiliriz.
`jpeg()` ile jpeg olarak kaydedebiliriz.
dev.off() işlevi plot işlemini bitirdiğimizi gösterir.

```{r chunk15}
pdf("Figure.pdf")
plot(x, y, col = "green")
dev.off()
```

 
seq() sayı dizisi oluşturmak için kullanılır.


```{r chunk16}
x <- seq(1, 10)
x
x <- 1:10
x
x <- seq(-pi, pi, length = 50)
```



```{r chunk17}
y = x

f = outer(x, y, function(x, y) cos(y) / (1 + x^2))

contour(x, y, f)

contour(x, y, f, nlevels = 45, add = T)

fa <- (f - t(f)) / 2

contour(x, y, fa, nlevels = 15)
?contour()



```


The `image()` function works the same way as `contour()`, except that it produces a color-coded plot whose colors depend on the `z` value. This is
 known as a , and is sometimes used to plot temperature in weather forecasts. Alternatively, `persp()` can be used to produce a three-dimensional plot.
 The arguments `theta` and `phi` control the angles at which the plot is viewed.

```{r chunk18}
image(x, y, fa)
persp(x, y, fa)
persp(x, y, fa, theta = 30)
persp(x, y, fa, theta = 30, phi = 20)
persp(x, y, fa, theta = 30, phi = 70)
persp(x, y, fa, theta = 30, phi = 40)
```


### Verileri İndeksleme 

Genellikle bir veri kümesinin bir bölümünü incelemek isteriz. Verilerimizin matriste saklandığını varsayalım A.

```{r chunk19}
A <- matrix(1:16, 4, 4)
A
```

Then, typing

```{r chunk20}
A[2, 3]

```

ikinci satıra ve üçüncü sütuna karşılık gelen öğeyi seçecektir.


Açık parantez simgesinden sonraki ilk sayı [ her zaman satırı, ikinci sayı ise her zaman sütunu ifade eder. 


İndeks olarak vektörler sağlayarak aynı anda birden fazla satır ve sütun da seçebiliriz.

```{r chunk21}
A[c(1, 3), c(2, 4)]
A[1:3, 2:4]
A[1:2, ]
A[, 1:2]
```

Son iki örnek, sütunlar için dizin veya satırlar için dizin içermez. 

RBunlar , sırasıyla tüm sütunları veya tüm satırları içermesi gerektiğini belirtir . 

R bir matrisin tek bir satırını veya sütununu bir vektör olarak ele alır.

```{r chunk22}
A[1, ]
```

The use of a negative sign `-` in the index tells `R` to keep all rows or columns except those indicated in the index.

```{r chunk23}
A[-c(1, 3), ]
A[-c(1, 3), -c(1, 3, 4)]
```


İşlev dim(), belirli bir matrisin satır sayısını ve ardından sütun sayısını verir.



```{r chunk24}
dim(A)
```


## Loading Data

Auto veri setimizi read.table() foonsiyonu ile okuya biliriz.
View veya head fonksiyonu ile verimizi görintüleyelim.


```{r chunk25}
Auto <- read.table("Auto.data")
View(Auto)
head(Auto)
```

Yukardaki tablo okuma kısmında bir problem var çünkü değişkenlerin isimlerinide tabloda ilk satıradahil ediyor.
haeder = T fonksiyonu  değişken isimlerini ilk satırdan kaldırır ve kolon isimlerine atar.

Ayrıca eksik verileride ? işareti olarak tabloda gösteriyor.
Bunun yerine " na.strings = ? " fonksiyonu ? işaretini nan değeri olarak atar.




```{r chunk26}
Auto <- read.table("Auto.data", header = T, na.strings = "?", stringsAsFactors = T)
View(Auto)
```

The `stringsAsFactors = T` argument tells `R` that any variable containing character strings should be interpreted as a qualitative variable, and that each distinct character string represents a distinct level for that qualitative variable.
An easy way to load  data from Excel into `R` is to save it as a csv (comma-separated values) file, and then use the `read.csv()` function.

```{r chunk27}
Auto <- read.csv("Auto.csv", na.strings = "?", stringsAsFactors = T)
View(Auto)
dim(Auto)
Auto[1:4, ]
```

The `dim()` function tells us that the data has $397$ observations, or rows, and nine variables, or columns. There are various ways to deal with the missing data.  In this case,   only five of the rows contain missing observations,
and  so we choose to use the `na.omit()` function to simply remove these rows.

```{r chunk28}
Auto <- na.omit(Auto)
dim(Auto)
```


Once the data are loaded correctly, we can use `names()` to check the variable names.

```{r chunk29}
names(Auto)
attach(Auto)
```


## Additional Graphical and Numerical Summaries

We can use the `plot()` function to produce *scatterplots*  of the quantitative variables. However, simply typing the variable names will produce an error message,
 because `R` does not know to look in the `Auto` data set for those variables.

```{r chunk30}
plot(cylinders, mpg)
```

To refer to a variable, we must type the data set and the variable name  joined with a \textR{\$} symbol.
 Alternatively, we can use the `attach()`  function in order to tell `R` to make the variables in this data frame available by name. 

```{r chunk31}
plot(Auto$cylinders, Auto$mpg)
attach(Auto)
plot(cylinders, mpg)
```

The `cylinders` variable is stored as a numeric vector, so `R` has treated it as quantitative. However, since there are only a small number of possible values for `cylinders`,
 one may prefer to treat it as a qualitative variable. The `as.factor()` function converts quantitative variables into qualitative variables.

```{r chunk32}
cylinders <- as.factor(cylinders)
cylinders
```

If the variable plotted on the $x$-axis is qualitative, then  *boxplots* will  automatically be produced
by the `plot()` function.  As usual, a number of options can be specified in order to customize the plots.

```{r chunk33}
plot(cylinders, mpg)
plot(cylinders, mpg, col = "red")
plot(cylinders, mpg, col = "red", varwidth = T)
plot(cylinders, mpg, col = "red", varwidth = T,
    horizontal = T)
plot(cylinders, mpg, col = "red", varwidth = T,
    xlab = "cylinders", ylab = "MPG")
```

The `hist()` function can be used to plot a . Note that `col = 2` has the same effect as `col = "red"`.

```{r chunk34}
hist(mpg)
hist(mpg, col = 2)
hist(mpg, col = 2, breaks = 15)
```

The `pairs()` function creates a *scatterplot matrix*, i.e. a scatterplot for every pair of variables. We can also produce scatterplots for just a subset of the variables.

```{r chunk35}
pairs(Auto)
pairs(
    ~ mpg + displacement + horsepower + weight + acceleration,
    data = Auto
  )
```


In conjunction with the `plot()` function,  `identify()` provides a useful interactive method for identifying the value of a particular variable for points on a plot. We pass in three arguments to `identify()`: the $x$-axis variable, the $y$-axis variable, and the variable whose values we would like to see printed for each point. Then clicking one or more  points in the plot and hitting Escape will cause `R` to print the values of the variable of interest. The numbers printed under the `identify()` function correspond to the rows for the selected points.

```{r chunk36}
plot(horsepower, mpg)
identify(horsepower, mpg)
```

The `summary()` function produces a numerical summary of each variable in a particular data set.

```{r chunk37}
summary(Auto)
```

For qualitative variables such as `name`,  `R` will list the number of observations that fall in each category. We can also produce a summary of just a single variable.

```{r chunk38}
summary(mpg)
```




Once we have finished using `R`, we type `q()` in order to shut it down, or quit.
When exiting `R`, we have the option to save the current  so that all
objects (such as data sets)
that we have created in this `R` session will be available next time.
 Before exiting `R`, we may want to save a record of all of the commands that we typed in the most recent session; this can be accomplished using the `savehistory()` function. Next time we enter `R`, we can load that history using
the `loadhistory()` function, if we wish. 


